---
title: "PRA2-TCVD: Cleaning"
author: "Javier Cantero Lorenzo"
date: '`r format(Sys.Date(),"%e de %B, %Y")`'
output:
  html_document:
    toc: yes
    number_sections: True
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries, include=FALSE}
library(VIM)
library(dplyr)
library(knitr)
library(stringr)
```


First we load the csv file, taking into account that the delimiter is a semicolon and the American standard of decimal separation is used. 

```{r read}
# load csv file
input_filename <- "../csv/estradiol_raw.csv"
dataset <- read.csv(input_filename, sep=";", dec=".")
```

To check if the information has been loaded correctly we check the number of records, the number of variables, and their name.

```{r explore}
# number of records
nrow(dataset)

# number of variables
ncol(dataset)

# name of the variables
colnames(dataset)
```

The characteristics of the dataset correspond to those expected, so we can continue with the analysis.

```{r}
kable(sample_n(dataset, 5))
```

Remove attribute X.

```{r}
dataset$X <- NULL
```

Select quantitative columns and force numerical class.

```{r}
# apply American standard of decimal separation
columns_quantitative <- c('Id', 'Entage', 'Numchild', 'Agefbo', 'Agemenar', 'Estradl', 'BMI', 'WHR')
dataset[columns_quantitative] <- sapply(dataset[columns_quantitative], gsub, pattern=',', replacement='.')

# convert to numerical data type
dataset[columns_quantitative] <- sapply(dataset[columns_quantitative], as.character)
dataset[columns_quantitative] <- sapply(dataset[columns_quantitative], as.numeric)
```

Identify and remove duplicate ID values.

```{r}
# duplicate values
subset(dataset$Id, duplicated(dataset$Id))

# remove rows with same id (keeping one)
dataset <- subset(dataset, !duplicated(dataset$Id))
```

Show unique Ethnic values and clean.
```{r}
kable(table(dataset$Ethnic))

# trim whitespaces
dataset$Ethnic <- trimws(dataset$Ethnic)

# title capitalization standard
dataset$Ethnic <- str_to_title(dataset$Ethnic)

# specific spelling errors
dataset$Ethnic <- str_replace_all(dataset$Ethnic, "Caucsian", "Caucasian")
dataset$Ethnic <- str_replace_all(dataset$Ethnic, "Caucacian", "Caucasian")
dataset$Ethnic <- str_replace_all(dataset$Ethnic, "Af Am", "African American")
```

Label Area and Anykids attributes.
```{r}
# data type transformation
dataset$Area <- as.factor(dataset$Area)

# one-hot encoding to label encoding
dataset$Area <- factor(dataset$Area, levels=c(0, 1), labels=c("Urban", "Rural"))
```

Identify outliers in attributes.
```{r}
boxplot.stats(dataset$Estradl)$out
boxplot.stats(dataset$Agemenar)$out
boxplot.stats(dataset$BMI)$out
boxplot.stats(dataset$WHR)$out

tmp_agefbo <- dataset$Agefbo
tmp_agefbo[tmp_agefbo==0] <- NA

boxplot.stats(tmp_agefbo)$out
```

Visualize boxplots.

```{r}
boxplot(dataset$Estradl, main='Estradl (with outliers)')
boxplot(dataset$Agemenar, main='Age of menarche')
boxplot(tmp_agefbo, main='Age of first born child')
boxplot(dataset$BMI, main='BMI')
boxplot(dataset$WHR, main='WHR')
```

Remove outliers.
```{r}
dataset$Estradl[dataset$Estradl > 100] <- NA
dataset$Agemenar[which(dataset$Agemenar %in% boxplot.stats(dataset$Agemenar)$out)] <- NA

dataset$Agefbo[which(dataset$Agefbo %in% boxplot.stats(dataset$Agefbo)$out)] <- NA
```

Inconsistencies Anykids vs Numchild
```{r}
# save indexes with NA values for Anykids attribute
anykids_na <- is.na(dataset$Anykids)

# create new Anykids attribute using Numchild
consistent_anykids <- as.numeric(dataset$Numchild > 0)

dataset$Anykids <- consistent_anykids
```

Inconsistencies Entage vs Agemenar
```{r}
# data type transformation
dataset$Anykids <- as.factor(dataset$Anykids)

# one-hot encoding to label encoding
dataset$Anykids <- factor(dataset$Anykids, levels=c(0, 1), labels=c("No", "Yes"))
```

Inconsistencies Agefbo vs Agemenar vs Numchild
```{r}
# temporal consistent attributes
tmp_entage <- pmax(dataset$Entage, dataset$Agemenar)
tmp_agemenar <- pmin(dataset$Entage, dataset$Agemenar)

dataset$Entage <- tmp_entage
dataset$Agemenar <- tmp_agemenar
```

Imputation of values
```{r}
# copy of original dataset before imputation
tmp_original_dataset <- dataset
ids_with_nans <- subset(dataset$Id, !complete.cases(dataset))

# temporal dataframes
tmp_dataset_afam <- subset(dataset, dataset$Ethnic == 'African American')
tmp_dataset_cauc <- subset(dataset, dataset$Ethnic == 'Caucasian')

# imputation with ethnic group neighbours
tmp_dataset_afam <- kNN(tmp_dataset_afam, variable=c('Agemenar', 'Agefbo', 'Entage', 'Estradl', 'Numchild'), k=3, imp_var=FALSE)

tmp_dataset_cauc <- kNN(tmp_dataset_cauc, variable=c('Agemenar', 'Agefbo', 'Entage', 'Estradl', 'Numchild'), k=3, imp_var=FALSE)

# join temporal datasets
dataset <- rbind(tmp_dataset_afam, tmp_dataset_cauc)
```

Show values before and after imputation.
```{r}
# original dataset
original_subdataset <- subset(tmp_original_dataset, tmp_original_dataset$Id %in% ids_with_nans)
kable(original_subdataset[order(original_subdataset$Id),])
```

```{r}
# imputed dataset
subdataset <- subset(dataset, dataset$Id %in% ids_with_nans)
kable(subdataset[order(subdataset$Id),])
```

Save cleaned result
```{r}
output_filename <- '../csv/estradiol_clean.csv'
write.csv(dataset, file=output_filename, row.names=FALSE)
```